---
title: "BSET tutorial"
author: "Pietro Carlotti"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

`BSET` (Bayesian Surrogate Evaluation Test) is an `R` package for assessing the validity of surrogate markers in clinical trials. It provides hypothesis testing tools to evaluate whether a surrogate can reliably estimate the causal effect of a treatment on a primary outcome. The package implements the imputation-based Bayesian methodology of Carlotti and Parast (2026), extending the frequentist rank-based approach of Parast et al. (2024). BSET addresses key limitations of the frequentist method, including the lack of causal interpretability and the inability to adjust for covariates in the estimation process.

The package supports Bayesian testing both with and without baseline covariates. Additionally, it includes comprehensive simulation suites to replicate studies from both papers, enabling performance comparisons between Bayesian and frequentist approaches across diverse clinical scenarios.

## Installation

The package is not yet available on CRAN, but you can install the development version of BSET from [GitHub](https://github.com/) with:

```{r install-BSET}
# install.packages("pak")
# pak::pak("PietroCarlotti/BSET")
```

## Generating the data

The package includes two functions to generate data for the simulations: `DGP` and `DGP_X`. The first one generates data from the simulation settings of Parast et al. (2025), which do not include covariates; the second one generates data which depends on a binary covariate \(X\) and is used for the simulations of Carlotti and Parast (2026).

```{r generate-data}

```
