---
title: "BSET Tutorial"
author: "Pietro Carlotti"
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    number_sections: false
    toc: true
vignette: >
  %\VignetteIndexEntry{BSET Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo=FALSE}
/* Make the overall page margins smaller / wider content area */
body {
  max-width: 900px !important; /* Default is usually around 700px; increase this for smaller margins */
  margin: auto;
}

/* Force the table to be wider */
table {
  width: 70%; /* Adjust this percentage as you like (e.g., 100% for full width) */
  margin-left: auto;
  margin-right: auto;
  margin-bottom: 20px;
}

/* Style the caption */
caption {
  caption-side: bottom !important;
  text-align: center;
  padding-top: 10px;
  width: 100%; /* Ensures the caption can use the full width of the table */
}
```

## Introduction

`BSET` (Bayesian Surrogate Evaluation Test) is an `R` package for assessing the validity of surrogate markers in clinical trials. It provides hypothesis testing tools to evaluate whether a surrogate can reliably estimate the causal effect of a treatment on a primary outcome. The package implements the imputation-based Bayesian methodology of Carlotti and Parast (2026), extending the frequentist rank-based approach of Parast et al. (2024). BSET addresses key limitations of the frequentist method, including the lack of causal interpretability and the inability to adjust for covariates in the estimation process.

The package supports Bayesian testing both with and without baseline covariates. Additionally, it includes comprehensive simulation suites to replicate studies from both papers, enabling performance comparisons between Bayesian and frequentist approaches across diverse clinical scenarios.

## Installation

The package is not yet available on CRAN, but you can install the development version of BSET from [GitHub](https://github.com/) with:

```{r install-BSET}
# install.packages("pak")
# pak::pak("PietroCarlotti/BSET")
```

## Generating the data

```{r set-seed}
# Set the random seed for reproducibility
set.seed(123)
```

The package includes two functions to generate data for the simulations: `DGP` and `DGP_X`. The first one generates data from the simulation settings of Parast et al. (2024), which do not include covariates; the second one generates data which depends on a binary covariate \(X\) and is used for the simulations of Carlotti and Parast (2026).

For example, this how we can generate data from the second setting defined in Parast et al. (2024) (i.e.: the setting where the surrogate is perfect and covariates are not included):

```{r generate-data-no-X}
# Sample size
n <- 50

# Treatment assignment probability
p <- 0.5

# Mean vector of potential outcomes for the primary outcome and the surrogate
mu_star <- c(6, 6, 2.5, 2.5)

# Covariance matrix of potential outcomes for the primary outcome and the surrogate
Sigma_star <- kronecker(
  diag(2),
  matrix(
    data = c(3, 3,
             3, 3.1),
    nrow = 2,
    ncol = 2)
)

# Generate the data
data_no_X <- BSET::DGP_no_X(
  n = n,
  p = p,
  mu_star = mu_star,
  Sigma_star <- Sigma_star,
  model = "Gaussian"
)
```

While this is how we can generate data from the first setting defined in Carlotti and Parast (2026) (i.e.: the setting where the surrogate is useful and covariates are included):

```{r generate-data-X}
# Sample size
n <- 50

# Treatment assignment probability
p <- 0.5

# Binary covariate probability
q <- 0.5

# Mean vector for potential outcomes when X = 0 and when X = 1
mu_0 <- c(5, 5, 0, 0)
mu_1 <- c(5, -5, 0, -10)

# Covariance matrix for potential outcomes when X = 0 and when X = 1
Sigma_0 <- kronecker(
  diag(2),
  matrix(
    data = c(1, 1,
             1, 2),
    nrow = 2,
    ncol = 2)
  )

Sigma_1 <- kronecker(
  diag(2),
  matrix(
    data = c(1, 1,
             1, 2),
    nrow = 2,
    ncol = 2)
  )

# Generate the data
data_X <- BSET::DGP_X_binary(
  n = n,
  p = p,
  q = q,
  mu_0 = mu_0,
  mu_1 = mu_1,
  Sigma_0 = Sigma_0,
  Sigma_1 = Sigma_1
  )
```

## Computing the true values of \(\delta\) and \(\theta\)

Given the data generated with the above functions, we can estimate \(\delta\) and \(\theta\) using the `compute_delta` and `compute_theta` functions, respectively.

For example, for the data generated without covariates:

```{r compute-delta-theta-no-X}
# Compute delta
delta_no_X <- BSET::compute_delta(MC_data = data_no_X)

# Compute theta
theta_no_X <- BSET::compute_theta(MC_data = data_no_X)
```

And for the data generated with covariates:

```{r compute-delta-theta-X}
# Compute delta
delta_X <- BSET::compute_delta(MC_data = data_X)

# Compute theta
theta_X <- BSET::compute_theta(MC_data = data_X)
```

The estimated values of \(\delta\) and \(\theta\) are shown in Table \@ref(tab:delta-theta-table).

```{r delta-theta-table, echo=FALSE}
delta_theta_table <- data.frame(
  Setting = c("No X (Perfect Surrogate)", "Binary X"),
  "$\\widehat{\\delta}$" = c(delta_no_X$delta, delta_X$delta),
  "$\\widehat{\\theta}$" = c(theta_no_X$theta, theta_X$theta),
  check.names = FALSE 
)

knitr::kable(
  delta_theta_table, 
  caption = "Estimated values of $\\delta$ and $\\theta$ in the two settings.",
  escape = FALSE,
  booktabs = TRUE,
  align = "ccc",
  digits = 3
)
```

To determine the values of \(\delta\) and \(\theta\) for all the settings defined in the two papers, we use a large-scale Monte Carlo simulation via the `compute_estimands_Parast_et_al_2024` function for data without covariates, and the `compute_estimands_Carlotti_and_Parast_2026` function for data including covariates.

```{r compute-delta-theta-true}
# Number of Monte Carlo samples to generate per setting
MC_samples <- 10#00000

# Compute the true values of delta and theta for all settings of Parast et al. (2024)
estimands_Parast_et_al_2024 <- BSET::compute_estimands_Parast_et_al_2024(MC_samples)

# Compute the true values of delta and theta for all settings of Carlotti and Parast (2026)
estimands_Carlotti_and_Parast_2026 <- BSET::compute_estimands_Carlotti_and_Parast_2026(MC_samples)
```

As shown in Table \@ref(tab:estimands-Parast-et-al-2024), \(\delta\) and \(\theta\) yield identical values (up to Monte Carlo error) across all settings. Both estimands are close to \(0\) for a perfect surrogate, they are significantly higher for a useless surrogate, and take intermediate values for imperfect surrogates or misspecified models.

```{r estimands-Parast-et-al-2024, echo=FALSE}
estimands_Parast_et_al_2024_table <- data.frame(
  Setting = c("Useless surrogate", "Perfect surrogate", "Imperfect surrogate", "Misspecified model"),
  "$\\delta$" = estimands_Parast_et_al_2024$delta_MC,
  "$\\theta$" = estimands_Parast_et_al_2024$theta_MC,
  check.names = FALSE
)

knitr::kable(
  estimands_Parast_et_al_2024_table,
  caption = "Monte Carlo estimates of $\\delta$ and $\\theta$ for the simulation settings considered in Parast et al. (2024).",
  booktabs = TRUE,
  align = "ccc",
  digits = 3
)
```

Whereas, as shown in Table \@ref(tab:estimands-Carlotti-and-Parast-2026), when we include covariates in the data generating process, \(\delta\) and \(\theta\) yield different values. In particular, while \(\theta\) is close to \(0\) for a perfect surrogate, \(\delta\) is significantly higher.

```{r estimands-Carlotti-and-Parast-2026, echo=FALSE}
estimands_Carlotti_and_Parast_2026_table <- data.frame(
  Setting = c("Perfect surrogate"),
  "$\\delta$" = estimands_Carlotti_and_Parast_2026$delta_MC,
  "$\\theta$" = estimands_Carlotti_and_Parast_2026$theta_MC,
  check.names = FALSE
)

knitr::kable(
  estimands_Carlotti_and_Parast_2026_table,
  caption = "Monte Carlo estimates of $\\delta$ and $\\theta$ for the simulation settings considered in Carlotti and Parast (2026).",
  booktabs = TRUE,
  align = "ccc",
  digits = 3
)
```

## Compute the validation threshold \(\eta\)



